Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

\textit{De novo} genes that do not provide any fitness benefit to an organism are not fixed in populations via natural selection. These genes can be lost due to mutation pressure. Some newborn \textit{de novo} genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way. These genes would thus be eliminated from the population genomes via negative selection. We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation). It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively). Furthermore, non-stop mutations can also lead to cellular toxicity. Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may be the better way to inactivate the associated genes.

@@

Previous studies have argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure. They based these conclusions on mutation rate bias estimated in bacteria. We asked if our model also makes similar predictions. To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic. We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (i.e A, C, F, I, L, M \& V). Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic. This finding is in agreement with the previous studies. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydropathy ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydropathic index values ($\alpha_x$).

>>>>

\textit{De novo} genes that do not provide any fitness benefit to an organism are not fixed in populations via natural selection. These genes can be lost due to mutation pressure. Some newborn \textit{de novo} genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way. These genes would thus be eliminated from the population genomes via negative selection. We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation). It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively). Furthermore, non-stop mutations can also lead to cellular toxicity \citep{nonstop}. Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may be the better way to inactivate the associated genes.

@@

A revious studies has argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure \citep{hydrophobicRatchet}. They based these conclusions on mutation rate bias estimated in bacteria \citep{bacteriaMutbias}. We asked if our model also makes similar predictions. To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic. We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (i.e A, C, F, I, L, M \& V). Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic. This finding is in agreement with the previous studies. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydropathy ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydropathic index values ($\alpha_x$).
